The need for accuracy and smoothness in numerical simulations

### 2024-05-12 report/src/draft.tex
Fixed misprints

\section{Motivation} Consider the problem resolving the motion of a system of atoms moving in a force-field $\bm{f}$ subject to a set of constraints $g$. In this case, Newton's 2nd law takes the form the following system of differential algebraic equations
\begin{align}
  \bm{q}'(t) &= \bm{v}(t) \\
  \bm{M}\bm{v}'(t) &= \bm{f}(\bm{q}(t)) - \bm{G}(\bm{q}(t))^T\bm{\lambda}(t) \\
  \bm{g}(\bm{q}(t)) &= \bm{0}. 
\end{align}
The vector $\bm{q}$ represents the position of the atoms.
The vector $\bm{v}$ represents the velocities of the atoms. The function $\bm{f}$ represents the force-field acting on the atoms.
The nonsingular diagonal matrix $\bm{M}$ lists the masses of the atoms.
The function $\bm{G}$ is the Jacobian of the constraint function $\bm{g}$ and $\bm{\lambda}$ is a vector of Lagrange multipliers.
In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm.
It uses a pair of staggered grids with uniform step size $h$ and takes the form
\begin{align}
  \bm{v}_{n+1/2} &= \bm{v}_{n-1/2} + \bm{h} \bm{M}^{-1} \left( \bm{f}(\bm{q}_n) - \bm{G}(\bm{q}_n)^T \bm{\lambda}_n \right), \\
  \bm{q}_{n+1} &= \bm{q}_n + h \bm{v}_{n + 1/2}, \\
  \bm{g}(\bm{q}_{n+1}) &= \bm{0}.
\end{align}
The constraint equation is a usually a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
Now let $T \in \mathbb{R}$ denote any target value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm. It is clear that $T$ and $A_h$ are both functions of the force-field $\bm{f}$ and the question of adjusting $\bm{f}$ to match the outcome of a physical experiment naturally suggests itself. Let therefore $T_0 \in \mathbb{R}$ be given and consider the problem of solving the equation
\begin{equation}
  T_0 = T(\bm{f})
\end{equation}
with respect to $\bm{f}$. The fundamental problem is that we cannot compute the exact values of $T(\bm{f})$ and $A_h(\bm{f})$. We must contend with the fact that we cannot solve the constraint equations exactly nor can we avoid rounding errors. Let $\hat{A}_h(\bm{f}, \tau, u)$ denote the value returned by our computer when solving the constraint equations with a relative error bounded by $\tau$ and using floating point arithmetic with unit roundoff $u$. Suppose that $T_0 - A_h(\bm{f}, \tau, u)$ is small. Can we show that $T_0 - T(\bm{f})$ is small? The triangle inequality implies the following inequality
\begin{equation}
  |T_0 - T(\bm{f}) | \leq  |T_0 - A_h(\bm{f},\tau,u)| + |A_h(\bm{f}) - A_h(\bm{f},\tau,u)| + |T(\bm{f}) - A_h(\bm{f})|.
\end{equation}
We conclude that $T_0 - T(\bm{f})$ is small, if the computational error $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the discretization error $T(\bm{f}) - A_h(\bm{f})$ are both small. If we cannot assert control over these errors, then we cannot say with certainty that our model delivers a good approximation of the physical reality.

@@

We mention in passing that low order methods are more practical than high order methods in the sense that low order methods tend to have asymptotic ranges that are larger than high order methods.
This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small. This issue is more acute for high order methods than for low order methods, because $A_h$ tends to $T$ more rapidly for high order methods than for low order methods.

@@

For each drag coefficient our target value $T$ is the maximum range of the shell as the elevation of the howitzer varies continuously from $0$ to $\frac{\pi}{2}$. For each drag coefficient, we compute $12$ different approximations $A_{h_k}$ of $T$ using the step size $h_k = 2^{3-k}$ seconds, where $k \in \{1,2,\dots,12\}$.
For each drag coefficient and for each value of the time step $h$, a range function is defined which returns the range of the shell as a function of the howitzers elevation $\theta$.
The range functions are unimodal and the maximum range is found using the golden section search algorithm.
The initial search bracket is $[0,\pi/2$] and this bracket is systematically reduced in length until it is shorter than $\frac{\pi}{2}u$.

The script will either read the raw data from a file or generate it from scratch. In any case, the script produces several figures and tables including Figure \ref{fig:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}. These two figures represent calculations where the final time step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff. Figure \ref{fig:maxrange_rk1_table_tol53} list the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$. In each case the error estimate suggests that the computed range is exact to number of figures shown. In particular, we see that a G7 type shell achieves a maximum range of 17.4 km and all other shells have ranges that are less than 16.9 km. It is however, a fallacy to conclude anything on the basis of this table alone. In each case, we need to assert that we are inside the asymptotic range and that the error estimates are reliable. To this end, we examine the evolution of Richardson's fraction for the maximum range of each shell, see Figure \ref{fig:maxrange_rk1_fraction_tol53}. For each of the 6 different drag coefficients we find that evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$. This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step. We observe that for each drag coefficient, $k=12$ is still inside the asymtotic range and we have no reason to doubt the magnitude of the error estimate. We conclude that the best model for the D-20 howitzer is in fact the G7 shell.

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=\linewidth]{maxrange_rk1_tol53.pdf}
  \caption{The evolution of Richardson's fraction corresponding to the maximum range of 6 different shells fired from the D-20 howitzer.} \label{fig:maxrange_rk1_fraction_tol53}
\end{figure}

@@

Firstly, the potential energy and especially the kinetic energy exhibits violent oscillations when the tolerance is large, i.e, $\tau = 10^{-4}$. The amplitude of the oscillations is reduced when $\tau = 10^{-12}$. We expect the solution of the underlying differential algebraic equation to behave nicely, but we have no such expection the computed approximation unless $\tau$ is very small.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{maxrange_rk2_tol25.pdf} \caption{The evolution of $F_h$ for the maximum range of 6 different shells computed using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-25}$.}
  \label{fig:maxrange_rk2_tol25}
\end{figure}

>>>>

\section{Motivation} Consider the problem resolving the motion of a system of atoms moving in a force-field $\bm{f}$ subject to a set of constraints $g$. In this case, Newton's 2nd law takes the form the following system of differential algebraic equations
\begin{align}
  \bm{q}'(t) &= \bm{v}(t) \\
  \bm{M}\bm{v}'(t) &= \bm{f}(\bm{q}(t)) - \bm{G}(\bm{q}(t))^T\bm{\lambda}(t) \\
  \bm{g}(\bm{q}(t)) &= \bm{0}. 
\end{align}
The vector $\bm{q}$ represents the position of the atoms.
The vector $\bm{v}$ represents the velocities of the atoms. The function $\bm{f}$ represents the force-field acting on the atoms.
The nonsingular diagonal matrix $\bm{M}$ lists the masses of the atoms.
The function $\bm{G}$ is the Jacobian of the constraint function $\bm{g}$ and $\bm{\lambda}$ is a vector of Lagrange multipliers.
In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm.
It uses a pair of staggered grids with uniform step size $h$ and takes the form
\begin{align}
  \bm{v}_{n+1/2} &= \bm{v}_{n-1/2} + \bm{h} \bm{M}^{-1} \left( \bm{f}(\bm{q}_n) - \bm{G}(\bm{q}_n)^T \bm{\lambda}_n \right), \\
  \bm{q}_{n+1} &= \bm{q}_n + h \bm{v}_{n + 1/2}, \\
  \bm{g}(\bm{q}_{n+1}) &= \bm{0}.
\end{align}
The constraint equation is a usually a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
Now let $T \in \mathbb{R}$ denote any target value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm. It is clear that $T$ and $A_h$ are both functions of the force-field $\bm{f}$ and the question of adjusting $\bm{f}$ to match the outcome of a physical experiment naturally suggests itself. Let therefore $T_0 \in \mathbb{R}$ be given and consider the problem of solving the equation
\begin{equation}
  T_0 = T(\bm{f})
\end{equation}
with respect to $\bm{f}$. The fundamental problem is that we cannot compute the exact values of $T(\bm{f})$ and $A_h(\bm{f})$. We must contend with the fact that we can hardly solve the constraint equations exactly nor can we avoid rounding errors. Let $\hat{A}_h(\bm{f}, \tau, u)$ denote the value returned by our computer when solving the constraint equations with a relative error bounded by $\tau$ and using floating point arithmetic with unit roundoff $u$. Suppose that $T_0 - A_h(\bm{f}, \tau, u)$ is small. Can we assert that $T_0 - T(\bm{f})$ is small? The triangle inequality implies the following inequality
\begin{equation}
  |T_0 - T(\bm{f}) | \leq  |T_0 - A_h(\bm{f},\tau,u)| + |A_h(\bm{f}) - A_h(\bm{f},\tau,u)| + |T(\bm{f}) - A_h(\bm{f})|.
\end{equation}
We conclude that $T_0 - T(\bm{f})$ is small, if the computational error $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the discretization error $T(\bm{f}) - A_h(\bm{f})$ are both small. If we cannot control these errors, then we cannot say with certainty that our model delivers a good approximation of the physical reality.

@@

We mention in passing that low order methods are more practical than high order methods in the sense that low order methods tend to have asymptotic ranges that are larger than high order methods.
This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small. This issue is more acute for high order methods than for low order methods, because $A_h$ tends to $T$ much more rapidly for high order methods than for low order methods.

@@

For each drag coefficient our target value $T$ is the maximum range of the shell as the elevation of the howitzer varies continuously from $0$ to $\frac{\pi}{2}$. For each drag coefficient, we compute $12$ different approximations $A_{h_k}$ of $T$ using the step size $h_k = 2^{3-k}$ seconds, where $k \in \{1,2,\dots,12\}$.
For each drag coefficient and for each value of the time step $h$, a range function is defined which returns the range of the shell as a function of the howitzers elevation $\theta$.
The range functions are unimodal and the maximum range is found using the golden section search algorithm.
The initial search bracket is $[0,\pi/2$] and this bracket is systematically reduced in length until it is shorter than $\frac{\pi}{2}u$.
The script will either read the raw data from a file or generate it from scratch. In any case, the script produces several figures and tables including Figure \ref{fig:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}. These two figures represent calculations where the final time step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff. Figure \ref{fig:maxrange_rk1_table_tol53} list the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$. In each case the error estimate suggests that the computed range is exact to number of figures shown. In particular, we see that a G7 type shell achieves a maximum range of 17.4 km and all other shells have ranges that are less than 16.9 km. It is however, a fallacy to conclude anything on the basis of this table alone. In each case, we need to assert that we are inside the asymptotic range and that the error estimates are reliable. To this end, we examine the evolution of Richardson's fraction for the maximum range of each shell, see Figure \ref{fig:maxrange_rk1_fraction_tol53}. For each of the 6 different drag coefficients we find that evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$. This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step. We observe that for each drag coefficient, $k=12$ is still inside the asymtotic range and we have no reason to doubt the magnitude of the error estimate. We conclude that the best model for the D-20 howitzer is in fact the G7 shell.

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=\linewidth]{maxrange_rk1_tol53.pdf}
  \caption{The evolution of Richardson's fraction corresponding to the maximum range of 3 different shells fired from the D-20 howitzer.} \label{fig:maxrange_rk1_fraction_tol53}
\end{figure}

@@

Firstly, the potential energy and especially the kinetic energy exhibits violent oscillations when the tolerance is large, i.e, $\tau = 10^{-4}$. The amplitude of the oscillations is reduced when $\tau = 10^{-12}$. We expect the solution of the underlying differential algebraic equation to behave nicely, but we have no such expectation for the computed approximation unless $\tau$ is very small.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{maxrange_rk2_tol25.pdf} \caption{The evolution of $F_h$ for the maximum range of 3 different shells computed using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-25}$.}
  \label{fig:maxrange_rk2_tol25}
\end{figure}
